---
title: "California Oil Well EDA"
author: "Lukas Lescano"
date: "2026-04-30"
format: pdf
---

# 1. Set up

## Packages

```{r}
#| label: packages
#| message: false

# general use and cleaning
library(tidyverse)
library(janitor)
library(snakecase)

# file organization
library(here)

# wrap axis labels
library(scales)

# reading .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)
```

# 2. Functions

```{r}
#| label: reading in data function

# Header is on row 2 in these CubeBrowser exports; row 1 is blank.
# Read everything as character first to avoid silent type coercion
read_cube <- function(path) {
  read_xlsx(path, skip = 1, col_types = "text",
            .name_repair = ~ ifelse(.x == "", "unknown_col", .x)) |>
    select(where(~ !all(is.na(.x))))
}

```

```{r}
#| label: inspect-fn

inspect_table <- function(df, name) {
  cat("==========================================\n")
  cat("TABLE:", name, "\n")
  cat("==========================================\n")
  cat("Dimensions:", nrow(df), "rows x", ncol(df), "cols\n\n")
  
  cat("--- Column names (raw) ---\n")
  print(names(df))
  
  cat("\n--- Per-column summary ---\n")
  summary_df <- tibble(
    column      = names(df),
    n_missing   = map_int(df, ~ sum(is.na(.x) | .x == "")),
    pct_missing = round(100 * n_missing / nrow(df), 1),
    n_unique    = map_int(df, ~ dplyr::n_distinct(.x, na.rm = TRUE)),
    example_vals = map_chr(df, ~ {
      vals <- unique(na.omit(.x))
      vals <- vals[vals != ""]
      paste(head(vals, 3), collapse = " | ")
    })
  )
  print(summary_df, n = Inf)
  
  cat("\n--- First 5 rows ---\n")
  print(head(df, 5))
  
  cat("\n--- Duplicate full rows:", sum(duplicated(df)), "---\n")
  
  invisible(summary_df)
}
```

```{r}
#| label: clean cube function

clean_cube <- function(df) {
  df |>
    rename("Resources (Million bbl)" = "unknown_col") |> 
    filter(!if_all(everything(), is.na)) |>   # drop fully blank separator rows
    fill(Project, Operator, `Supply Segment Group`, 
         `Life Cycle Category`, Longitude, Latitude,
         Country, `US State`, Year, Basin, .direction = "down") |>  # fill ALL group key columns
    clean_names()
}
```

# 3. Data

## Read in data

```{r}
#| label: data-raw
#| message: false

table1_raw <- read_cube(here("data", "CubeBrowserTableExport_2322B09F.xlsx"))
table2_raw <- read_cube(here("data", "CubeBrowserTableExport_75677428.xlsx"))
table3_raw <- read_cube(here("data", "CubeBrowserTableExport_41525648.xlsx"))
```

# 4. Exploring Data Structure

## Inspeting data

```{r}
#| label: inspect-all

s1 <- inspect_table(table1_raw, "table1 (2322B09F)")
s2 <- inspect_table(table2_raw, "table2 (75677428)")
s3 <- inspect_table(table3_raw, "table3 (41525648)")
```

## Differences between tables

```{r}
#| label: checking for differences

# Compare column names across all three
names(table1_raw)
names(table2_raw)
names(table3_raw)

# Are they exactly the same?
identical(names(table1_raw), names(table2_raw))
identical(names(table1_raw), names(table3_raw))

# If not identical, find the differences
setdiff(names(table1_raw), names(table2_raw))
setdiff(names(table2_raw), names(table1_raw))
```

## Duplicates in tables

```{r}
#| label: duplicates

combined_raw <- bind_rows(
  table1_raw |> mutate(source_file = "2322B09F"),
  table2_raw |> mutate(source_file = "75677428"),
  table3_raw |> mutate(source_file = "41525648")
)

# How many full duplicate rows exist across files?
get_dupes(combined_raw)

# How many duplicates in table 1?
get_dupes(table1_raw)
```

## Exploring combined data structure

```{r}
#| label: exploring combined data structure 
#| 
# Check every column: unique values for categoricals, range for numerics
combined_raw |> count(source_file)
combined_raw |> count(supply_segment_group)
combined_raw |> count(life_cycle_category)
combined_raw |> count(opec_plus)
combined_raw |> count(gas_market)
combined_raw |> count(reservoir_depth_detail)
combined_raw |> count(country)
combined_raw |> count(us_state)
combined_raw |> count(basin)
combined_raw |> count(year)

# For the bucketed text columns
combined_raw |> count(internal_rate_of_return) |> arrange(desc(n)) |> print(n = Inf)
combined_raw |> count(payback_years) |> arrange(desc(n)) |> print(n = Inf)

# For numeric columns, check range and whether any non-numeric strings snuck in
combined |>
  summarise(across(c(longitude, latitude, eur_per_well, capex_per_eur,
                     capex_per_well, hyperbolic_factor_b_per_well,
                     breakeven_oil_price, breakeven_gas_price20,
                     breakeven_gas_price15, x24h_initial_production_per_well,
                     resources_million_bbl),
                   ~ sum(is.na(suppressWarnings(as.numeric(.x)))),
                   .names = "non_numeric_{.col}"))
```


## Unique values for combined

```{r}
#| label: unique values for each column

map(combined |> select(internal_rate_of_return, payback_years, 
                              resources_million_bbl, project, operator,
                              supply_segment_group, life_cycle_category,
                              basin, us_state, gas_market, 
                              reservoir_depth_detail, opec_plus,
                              well_breakeven_oil_price_detail,
                              breakeven_oil_price_detail,
                              breakeven_gas_price15_detail),
    ~ sort(unique(.x)))
```


# 5. Wrangling and Cleaning data

## Clean cube 

```{r}
#| label: clean cube

table1_clean <- clean_cube(table1_raw)
table2_clean <- clean_cube(table2_raw)
table3_clean <- clean_cube(table3_raw)
```

## Combined data cleaning

```{r}
#| label: creating combined clean data frame

combined_clean <- combined |>
  mutate(
    # Numeric conversions
    across(c(longitude, latitude, eur_per_well, capex_per_eur,
             capex_per_well, hyperbolic_factor_b_per_well,
             breakeven_oil_price, breakeven_gas_price20,
             breakeven_gas_price15, x24h_initial_production_per_well),
           as.numeric),
    year = as.integer(year),
    
    # Factor conversions
    supply_segment_group   = factor(supply_segment_group),
    life_cycle_category    = factor(life_cycle_category),
    basin                  = factor(basin),
    us_state               = factor(us_state),
    gas_market             = factor(gas_market),
    source_file            = factor(source_file),
    reservoir_depth_detail = factor(reservoir_depth_detail,
      levels = c("0-1000 m depth", "1000-2000 m depth", "2000-3000 m depth",
                 "3000-4000 m depth", "4000-5000 m depth", "5000-6000 m depth"),
      ordered = TRUE)
  )
```


***NEED TO ADDRESS IRR, and payback years, and resources_million_bbl***

# 6. IRR Profitability Figure

Grouping/categorical context:
- *basin*: geographic grouping
- *operator*: who is operating well
- *project*: which project is it
- *life_cycle_category*: producing vs discovery
- *supply_segment_group*: Shale vs. Onshore vs. Offshore
- *us_state*: which state

Economic context that pairs with IRR:
**breakeven_oil_price**: directly related to profitability
**capex_per_well**: cost side of the equation
**payback_years**: complementary profitability indicator
**resources_million_bbl**: size of the opportunity

Possible use:
- *reservoir_depth_detail*: cost driver that influences IRR
- *eur_per_well*: production side

